Early environmental conditions do not impact behavioural flexibility in an invasive and non-invasive lizard species

### 2025-01-12 ms.qmd
updates

^1^ Division of Ecology and Evolution, Research School of Biology, The Australian National University, Canberra, ACT 2601, Australia  
^2^ Flinder's University, College of Science and Engineering, Bedford Park, SA 5042, Australia  
^3^ Melbourne University, Melbourne, VIC 3000, Australia  
^4^ University of Wollongong, Wollongong, NSW 2500, Australia  
^5^ Université de Lille, Lille 59000, France  
$\ddagger$ Corresponding author: pablo.reciosantiago@anu.edu.au

@@

Here, we studied the effects of prenatal temperature and corticosterone (CORT) (the main GC in reptiles) [@Crino_2023], on behavioural flexibility in two species of reptiles, the delicate skink (_Lampropholis delicata_) and the common garden skink (_L. guichenoti_). Both species have similar life-histories and overlapping distributions [@chapple_know_2011; @chapple_biology_2014], but only _Lampropholis delicata_ has been successful in colonising different areas around the globe [@baker1979rainbow; @chapple2013divergent; @chapple2015deliinvLHI]. Previous studies exploring behavioural differences between the two species have found _L. delicata_ to be more exploratory than _L. guichenoti_ [@chapple_know_2011], but no differences in learning were observed between the skinks in an associative learning task [@bezzina2014does]. Disparities in behavioural flexibility may be driving the differences in invasion success between both skinks, a prediction supported in other invasive species [@sol2000behavioural; @sol2008brain; @wright2010behavioral; @amiel_smart_2011; @chapple2012can]. However, the relative impact of early environments on behavioural flexibility remains unexplored. Early environments are known to affect suites of morphological and behavioural traits differently in each species [@carrasco2023maternal; @kar2023heritability]. We manipulated CORT and incubation temperature in both species and then subjected hatchlings to a reversal learning task. Reversal learning is a widely employed tool to measure behavioural flexibility [@gapp2014early; @hurtubise2017effects], as it assesses an individual's ability to reverse a previously learnt behaviour providing researchers with a clear indicator of their ability to adjust to new conditions [@brown2010behavioral].

@@

#### Statistical analyses  
We performed the analyses for each species separately. We fitted two different Bayesian multilevel models using the brm function from the _brms_ package [@burkner2017brms] using an R extension (version 2.8.2) [@R]. We ran four parallel MCMC chains of 3000 iterations for each model, with a warmup period of 1000 iterations. We modelled correct choices [correct (1) or not (0)] as the response variable, and trial, hormone (CORT versus Control), incubation temperature (Cold versus Hot), along with the three-way interaction between trial, hormone, and temperature. If early environments impact learning then we would predict that the rate of learning (i.e., trial slope) varies by treatment as captured by the interactions. The error structure was modelled using a Bernoulli distribution with a logit link function (family = Bernoulli(link = 'logit')). We included a random intercept and slope (trial) for each lizard in our models. We also incorporated the clutch identity as a random factor. _L. delicata_ lays one clutch per year, while _L. guichenoti_ lays two [@chapple_know_2011; @chapple_biology_2014]. Since eggs were collected during half of the breeding season, clutches likely come from different mothers. Additionally, previous research has shown that clutches are generally sired by a single male, but sperm storage can occur [@kar2023heritability]. Given our partial split-clutch design, and the fact that maternal effects are expected to be stronger than paternal effects in these species, including the clutch as a random factor should account for the effects of parental identity.

@@

```{r, fig-rawdata}
#| label: fig-rawdata
#| fig.cap: "Probability of choosing correctly over trials for each of the treatments and species. The dots represent the proportion of individuals per trial that chose correctly while the lines show the mean predicted probability of choosing correctly, estimated using the posteriors of our model."
#
# Load the data
data_raw <- clean_df %>%
  group_by(species, trt, trial_reversal) %>%
  summarize(prop_correct = mean(FC_reversal, na.rm = TRUE)) %>%
  mutate(trt = factor(trt, 
          levels = c("A_23", "B_23", "A_28", "B_28"), 
          labels = c("CORT-Cold", "Control-Cold", "CORT-Hot", "Control-Hot")),
        species = factor(species,
          levels = c("delicata", "guichenoti"),
          labels = c("italic('L. delicata')", "italic('L. guichenoti')"))) %>%
  rename(Trial = trial_reversal,
         Treatment = trt) %>%
data.frame()
#
# Get the df for the predicted prob
df_deliprob$species <- "italic('L. delicata')"
df_guichprob$species <- "italic('L. guichenoti')"
df_combined <- rbind(df_deliprob, df_guichprob)
df_all <- full_join(df_combined, data_raw, by = c("Trial", "Treatment", "species"))
# Plot the raw data for the reversal tasks for both species with the predicted prob
plot_raw <- ggplot(data = df_all, aes(x = Trial, y = Mean_Predicted_prob, color = Treatment, fill = Treatment)) +
  geom_smooth(linewidth = 1) +
  geom_point(data = df_all, aes(x = Trial, y = prop_correct, color = Treatment, fill = Treatment)) +
  scale_color_manual(values = c("darkblue", "#68bde1", "#b50101", "#fa927d")) +
  scale_fill_manual(values = c("darkblue", "#68bde1", "#b50101", "#fa927d")) +
  facet_grid(species ~ Treatment, labeller = label_parsed, switch = "y") +
  labs(x = "Trial", y = "Proportion of correct choices", color = "Treatments", fill = "Treatments") +
  theme_classic()
#
# Save the plot
ggsave("./output/figures/plot_raw.png", plot=plot_raw, width = 25, height = 18, units = "cm", dpi = 600)
knitr::include_graphics("./output/figures/plot_raw.png")
```

@@

Plots for the different models of the reversal task:  
1.- _L. delicata_    
```{r , plotmod_drr, out.width="70%", fig.align="center"}
bayes_R2(mod_deli)
plot(mod_deli)
```

@@

2.- _L. guichenoti_  
```{r , plotmod_grr, out.width="70%", fig.align="center"}
bayes_R2(mod_guich)
plot(mod_guich)
```

@@

```{r, fig-age} 
#| label: fig-age
#| fig.cap: "Distribution of the age of the lizards by treatment and species"
# Plotting the distribution of the age of the lizards by treatment and species
# Load the data
data_age <- clean_df
# Plot the distribution of the age of the lizards by treatment and species
img_deli <- readPNG("./Others/Deli.png")
plot_age_deli <- data_age %>%
  filter(species == "delicata") %>%
  mutate(trt = factor(trt, 
          levels = c("A_23", "B_23", "A_28", "B_28"), 
          labels = c("Control-Cold", "CORT-Cold", "Control-Hot", "CORT-Hot"))) %>%
  ggplot(aes(x = age.start, color = trt)) +
  geom_density(linewidth = 1.5) +
  scale_color_manual(values = c("CORT-Cold"="darkblue", "Control-Cold"="#68bde1", "CORT-Hot"="#b50101", "Control-Hot"="#fa927d")) +  
  labs(title = "L. delicata", x = "Age (days)", color = "Treatments") +
  theme_classic() +
  theme(plot.margin = margin(5.5, 5.5, 5.5, 5.5, "mm"),
    plot.title = element_text(size = 14, family = "Times", face = "italic"),
    axis.text.y = element_blank(),   # Remove y-axis labels
    axis.title = element_text(size = 12, family = "Times"),
    axis.text = element_text(size = 10, family = "Times"),
    legend.position = "right",
    legend.title = element_text(size = 12, family = "Times"),
    legend.text = element_text(size = 11, family = "Times")
    ) +
  annotation_custom(rasterGrob(img_deli), xmin = 0.73, xmax = 0.98, ymin = 0.73, ymax = 0.98)
#
img_guich <- readPNG("./Others/Guich.png")
plot_age_guich <- data_age %>%
  filter(species == "guichenoti") %>%
  mutate(trt = factor(trt, 
          levels = c("A_23", "B_23", "A_28", "B_28"), 
          labels = c("Control-Cold", "CORT-Cold", "Control-Hot", "CORT-Hot"))) %>%
  ggplot(aes(x = age.start, color = trt)) +
  geom_density(linewidth = 1.5) +
  scale_color_manual(values = c("CORT-Cold"="darkblue", "Control-Cold"="#68bde1", "CORT-Hot"="#b50101", "Control-Hot"="#fa927d")) +  
  labs(title = "L. guichenoti", x = "Age (days)", color = "Treatments") +
  theme_classic() +
  theme(plot.margin = margin(5.5, 5.5, 5.5, 5.5, "mm"),
    plot.title = element_text(size = 14, family = "Times", face = "italic"),
    axis.text.y = element_blank(),   # Remove y-axis labels
    axis.title = element_text(size = 12, family = "Times"),
    axis.text = element_text(size = 10, family = "Times"),
    legend.position = "right",
    legend.title = element_text(size = 12, family = "Times"),
    legend.text = element_text(size = 11, family = "Times")
    ) +
  annotation_custom(rasterGrob(img_guich), xmin = 0.73, xmax = 0.98, ymin = 0.73, ymax = 0.98)
#
plot_age <- plot_grid(plot_age_deli, plot_age_guich, ncol = 1)
ggsave("./output/figures/plot_age.png", plot=plot_age, width = 25, height = 18, units = "cm", dpi = 600)
knitr::include_graphics("./output/figures/plot_age.png")
```

>>>>

^1^ Division of Ecology and Evolution, Research School of Biology, The Australian National University, Canberra, ACT 2601, Australia  
^2^ College of Science and Engineering, Flinders University, Bedford Park, SA 5001, Australia  
^3^ School of BioSciences, The University of Melbourne, Melbourne, VIC 3000, Australia  
^4^ School of Science, University of Wollongong, Wollongong, NSW 2500, Australia  
^5^ Université de Lille, Lille 59000, France  
$\ddagger$ Corresponding author: pablo.reciosantiago@anu.edu.au

@@

Here, we studied the effects of prenatal temperature and corticosterone (CORT) (the main GC in reptiles) [@Crino_2023], on behavioural flexibility in two species of reptiles, the delicate skink (_Lampropholis delicata_) and the common garden skink (_L. guichenoti_). Both species have similar life-histories and overlapping distributions [@chapple_know_2011; @chapple2015deliinvLHI], but only _Lampropholis delicata_ has been successful in colonising different areas around the globe [@baker1979rainbow; @chapple2013divergent; @chapple2015deliinvLHI]. Previous studies exploring behavioural differences between the two species have found _L. delicata_ to be more exploratory than _L. guichenoti_ [@chapple_know_2011], but no differences in learning were observed between the skinks in an associative learning task [@bezzina2014does]. Disparities in behavioural flexibility may be driving the differences in invasion success between both skinks, a prediction supported in other invasive species [@sol2000behavioural; @sol2008brain; @wright2010behavioral; @amiel_smart_2011; @chapple2012can]. However, the relative impact of early environments on behavioural flexibility remains unexplored. Early environments are known to affect suites of morphological and behavioural traits differently in each species [@carrasco2023maternal; @kar2023heritability]. We manipulated CORT and incubation temperature in both species and then subjected hatchlings to a reversal learning task. Reversal learning is a widely employed tool to measure behavioural flexibility [@gapp2014early; @hurtubise2017effects], as it assesses an individual's ability to reverse a previously learnt behaviour providing researchers with a clear indicator of their ability to adjust to new conditions [@brown2010behavioral].

@@

#### Statistical analyses  
We performed the analyses for each species separately. We fitted two different Bayesian multilevel models using the brm function from the _brms_ package [@burkner2017brms] using an R extension (version 2.8.2) [@R]. We ran four parallel MCMC chains of 3000 iterations for each model, with a warmup period of 1000 iterations. We modelled correct choices [correct (1) or not (0)] as the response variable, and trial, hormone (CORT versus Control), incubation temperature (Cold versus Hot), along with the three-way interaction between trial, hormone, and temperature. If early environments impact learning then we would predict that the rate of learning (i.e., trial slope) varies by treatment as captured by the interactions. The error structure was modelled using a Bernoulli distribution with a logit link function (family = Bernoulli(link = 'logit')). We included a random intercept and slope (trial) for each lizard in our models. We also incorporated the clutch identity as a random factor. _L. delicata_ lays one clutch per year, while _L. guichenoti_ lays two [@chapple_know_2011; @chapple2015deliinvLHI]. Since eggs were collected during half of the breeding season, clutches likely come from different mothers. Additionally, previous research has shown that clutches are generally sired by a single male, but sperm storage can occur [@kar2023heritability]. Given our partial split-clutch design, and the fact that maternal effects are expected to be stronger than paternal effects in these species, including the clutch as a random factor should account for the effects of parental identity.

@@

```{r, fig-rawdata}
#| label: fig-rawdata
#| fig.cap: "Probability of choosing correctly over trials for each of the treatments and species. The dots represent the proportion of individuals per trial that chose correctly while the lines show the mean predicted probability of choosing correctly, estimated using the posteriors of our model."
#
# Load the data
data_raw <- clean_df %>%
  group_by(species, trt, trial_reversal) %>%
  summarize(prop_correct = mean(FC_reversal, na.rm = TRUE)) %>%
  mutate(trt = factor(trt, 
          levels = c("A_23", "B_23", "A_28", "B_28"), 
          labels = c("CORT-Cold", "Control-Cold", "CORT-Hot", "Control-Hot")),
        species = factor(species,
          levels = c("delicata", "guichenoti"),
          labels = c("italic('L. delicata')", "italic('L. guichenoti')"))) %>%
  rename(Trial = trial_reversal,
         Treatment = trt) %>%
data.frame()
#
# Get the df for the predicted prob
df_deliprob$species <- "italic('L. delicata')"
df_guichprob$species <- "italic('L. guichenoti')"
df_combined <- rbind(df_deliprob, df_guichprob)
df_all <- full_join(df_combined, data_raw, by = c("Trial", "Treatment", "species"))
# Plot the raw data for the reversal tasks for both species with the predicted prob
plot_raw <- ggplot(data = df_all, aes(x = Trial, y = Mean_Predicted_prob, color = Treatment, fill = Treatment)) +
  geom_smooth(linewidth = 1) +
  geom_point(data = df_all, aes(x = Trial, y = prop_correct, color = Treatment, fill = Treatment)) +
  scale_color_manual(values = c("darkblue", "#68bde1", "#b50101", "#fa927d")) +
  scale_fill_manual(values = c("darkblue", "#68bde1", "#b50101", "#fa927d")) +
  facet_grid(species ~ Treatment, labeller = label_parsed, switch = "y") +
  labs(x = "Trial", y = "Proportion of correct choices", color = "Treatments", fill = "Treatments") +
  theme_classic()
#
# Save the plot
ggsave("./output/Figs_supplementary/Figure_S1.png", plot=plot_raw, width = 25, height = 18, units = "cm", dpi = 600)
knitr::include_graphics("./output/Figs_supplementary/Figure_S1.png")
```

@@

Plots for the different models of the reversal task:  
1.- _L. delicata_    
```{r , plotmod_drr, out.width="70%", fig.align="center"}
bayes_R2(mod_deli)
plot_mod_deli_list <- plot(mod_deli)
#
plot_mod_deli_1 <- plot_mod_deli_list[[1]]
plot_mod_deli_2 <- plot_mod_deli_list[[2]]
plot_mod_deli_3 <- plot_mod_deli_list[[3]]
#
plot_mod_deli <- plot_grid(plot_mod_deli_1, plot_mod_deli_2, plot_mod_deli_3, ncol = 1)
ggsave("./output/Figs_supplementary/Figure_S2.png", plot=plot_mod_deli, width = 20, height = 35, units = "cm", dpi = 600, bg = "white")
knitr::include_graphics("./output/Figs_supplementary/Figure_S2.png")
```

@@

2.- _L. guichenoti_  
```{r , plotmod_grr, out.width="70%", fig.align="center"}
bayes_R2(mod_guich)
plot_mod_guich_list <- plot(mod_guich)
#
plot_mod_guich_1 <- plot_mod_guich_list[[1]]
plot_mod_guich_2 <- plot_mod_guich_list[[2]]
plot_mod_guich_3 <- plot_mod_guich_list[[3]]
#
plot_mod_guich <- plot_grid(plot_mod_guich_1, plot_mod_guich_2, plot_mod_guich_3, ncol = 1)
ggsave("./output/Figs_supplementary/Figure_S3.png", plot=plot_mod_guich, width = 20, height = 35, units = "cm", dpi = 600, bg = "white")
knitr::include_graphics("./output/Figs_supplementary/Figure_S3.png")
```

@@

```{r, fig-age} 
#| label: fig-age
#| fig.cap: "Distribution of the age of the lizards by treatment and species"
# Plotting the distribution of the age of the lizards by treatment and species
# Load the data
data_age <- clean_df
# Plot the distribution of the age of the lizards by treatment and species
img_deli <- readPNG("./Others/Deli.png")
plot_age_deli <- data_age %>%
  filter(species == "delicata") %>%
  mutate(trt = factor(trt, 
          levels = c("A_23", "B_23", "A_28", "B_28"), 
          labels = c("Control-Cold", "CORT-Cold", "Control-Hot", "CORT-Hot"))) %>%
  ggplot(aes(x = age.start, color = trt)) +
  geom_density(linewidth = 1.5) +
  scale_color_manual(values = c("CORT-Cold"="darkblue", "Control-Cold"="#68bde1", "CORT-Hot"="#b50101", "Control-Hot"="#fa927d")) +  
  labs(title = "L. delicata", x = "Age (days)", color = "Treatments") +
  theme_classic() +
  theme(plot.margin = margin(5.5, 5.5, 5.5, 5.5, "mm"),
    plot.title = element_text(size = 14, family = "Times", face = "italic"),
    axis.text.y = element_blank(),   # Remove y-axis labels
    axis.title = element_text(size = 12, family = "Times"),
    axis.text = element_text(size = 10, family = "Times"),
    legend.position = "right",
    legend.title = element_text(size = 12, family = "Times"),
    legend.text = element_text(size = 11, family = "Times")
    ) +
  annotation_custom(rasterGrob(img_deli), xmin = 0.73, xmax = 0.98, ymin = 0.73, ymax = 0.98)
#
img_guich <- readPNG("./Others/Guich.png")
plot_age_guich <- data_age %>%
  filter(species == "guichenoti") %>%
  mutate(trt = factor(trt, 
          levels = c("A_23", "B_23", "A_28", "B_28"), 
          labels = c("Control-Cold", "CORT-Cold", "Control-Hot", "CORT-Hot"))) %>%
  ggplot(aes(x = age.start, color = trt)) +
  geom_density(linewidth = 1.5) +
  scale_color_manual(values = c("CORT-Cold"="darkblue", "Control-Cold"="#68bde1", "CORT-Hot"="#b50101", "Control-Hot"="#fa927d")) +  
  labs(title = "L. guichenoti", x = "Age (days)", color = "Treatments") +
  theme_classic() +
  theme(plot.margin = margin(5.5, 5.5, 5.5, 5.5, "mm"),
    plot.title = element_text(size = 14, family = "Times", face = "italic"),
    axis.text.y = element_blank(),   # Remove y-axis labels
    axis.title = element_text(size = 12, family = "Times"),
    axis.text = element_text(size = 10, family = "Times"),
    legend.position = "right",
    legend.title = element_text(size = 12, family = "Times"),
    legend.text = element_text(size = 11, family = "Times")
    ) +
  annotation_custom(rasterGrob(img_guich), xmin = 0.73, xmax = 0.98, ymin = 0.73, ymax = 0.98)
#
plot_age <- plot_grid(plot_age_deli, plot_age_guich, ncol = 1)
ggsave("./output/Figs_supplementary/Figure_S4.png", plot=plot_age, width = 25, height = 18, units = "cm", dpi = 600)
knitr::include_graphics("./output/Figs_supplementary/Figure_S4.png")
```
